---
title: "Assignment 8: Time Series Analysis"
author: "Jaclyn Narleski"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}

# Load libraries and check working directory
library(tidyverse)
library(lubridate)
library(zoo)
library(trend)
library(here)
here()

# Create ggplot theme
turqtheme <- theme(
  panel.background = element_rect(fill = "ivory"),
  panel.border = element_rect(color = "turquoise4", fill = NA, linetype = 1),
  plot.title = element_text(hjust = 0.5, size = 12, color = "turquoise4", 
    face = "bold"),
  axis.title = element_text(size = 10, color = "turquoise4"),
  panel.grid.major = element_line(
    color = "turquoise4", linetype = 2, size = 0.25),
  panel.grid.minor = element_line(
    color = "turquoise3", linetype = 3, size = 0.25),
  axis.text = element_text(color = "turquoise4"),
  axis.ticks = element_line(color = "turquoise4"),
  legend.title = element_text(size = 8, color = "turquoise4"),
  legend.position = "right")

theme_set(turqtheme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1

# Read csvs
O3.2010 <- read.csv(here(
  "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv"))
O3.2011 <- read.csv(here(
  "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv"))
O3.2012 <- read.csv(here(
  "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv"))
O3.2013 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv"))
O3.2014 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv"))
O3.2015 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv"))
O3.2016 <- read.csv(here( "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv"))
O3.2017 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv"))
O3.2018 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv"))
O3.2019 <- read.csv(here(
   "~/EDE_Fall2024/Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv"))

# Combine data frames into one
OzoneCombined <- 
  bind_rows(O3.2010, O3.2011, O3.2012, O3.2013, O3.2014, O3.2015, O3.2016, 
            O3.2017, O3.2018, O3.2019)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3

# Make date characters into Date class
OzoneCombined$Date <- as.Date(OzoneCombined$Date, format = "%m/%d/%Y")

# 4

# Wrangle data set using select()
OzoneCombined.Condensed <- OzoneCombined %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

# 5

# Generate data frame of dates by day
Days <- as.data.frame(seq(as.Date("2010-01-01"), as.Date("2019-12-31"), 
                          by = "day"))
# rename column name
colnames(Days) <- "Date"
  
# 6

# Combine Days data frame and ozone data 
GaringerOzone <- left_join(Days, OzoneCombined.Condensed, by = "Date") 

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7

# Create ggplot
OzoneOverTime <- GaringerOzone %>%
  ggplot(aes(
    x = Date,
    y = Daily.Max.8.hour.Ozone.Concentration)
  ) +
  geom_line(color = "chartreuse3") +
  geom_smooth(method = "lm", color = "deepskyblue4") +
  ggtitle("Ozone Concentrations Over Time") +
  ylab("Ozone Concentration (ppm)") +
  xlab("Year") 

# View plot
OzoneOverTime
```

>Answer:

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station?
- Ozone concentrations have decreased over time at this station.

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8

# Fill missing daily data using a linear interpolation
GaringerOzone$Daily.Max.8.hour.Ozone.Concentration <- na.approx(
  GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, na.rm = FALSE)

```

> Answer: Linear interpolation is best to fill in missing data for daily ozone concentrations because it estimates missing values based on avergae values that do not flunctuate extremely over time. Piecewise constant and spline interpolation anticipate more sudden changes and flunctuations in data. 


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9

# Generate new data frame of monthly mean ozone concentrations
GaringerOzone.monthly <- GaringerOzone %>%
  mutate(
    Year = year(Date),
    Month = month(Date)
  ) %>%
  group_by(Year, Month) %>%
  summarize(
    meanOzone = mean(Daily.Max.8.hour.Ozone.Concentration, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  # create date column set to first day of month for graphing purposes
  mutate(Date = my(paste(Month, "-", Year)))

```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10

# Generate time series for daily values
GaringerOzone.daily.ts <- ts(
  GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, frequency = 365,
  start = c(2010, 1))

# Generate time series for monthly values
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$meanOzone, frequency = 12,
                               start = c(2010, 1))
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11

# Decompose daily time series object
GaringerOzone.daily.decomposed <- stl(GaringerOzone.daily.ts, 
                                      s.window = "periodic")

# Plot
plot(GaringerOzone.daily.decomposed)

# Decompose monthly time series object 
GaringerOzone.monthly.decomposed <- stl(GaringerOzone.monthly.ts,
                                        s.window = "periodic")

# Plot
plot(GaringerOzone.monthly.decomposed)

```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12

# Run a seasonal Mann-Kendall monotonic trend analysis on monthly ozone 
MonthlyOzone.trend <- trend::smk.test(GaringerOzone.monthly.ts)

# Show results
MonthlyOzone.trend
summary(MonthlyOzone.trend)

```

> Answer: The seasonal Mann-Kendall test is appropriate because it accounts for seasonal variations (ozone tends to increase in the summer due to increase in solar radiation) and it doesn't assume that the data has a normal or specific distribution as a non-parametric statistical test. This helps the test account for outliers or missing values in the data set.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13

# Create ggplot
MonthlyOzone.plot <- ggplot(GaringerOzone.monthly, 
                            aes(x = Date,y = meanOzone)
  ) + 
  geom_point(color = "turquoise4") +
  geom_line(aes(group = 1), color = "chartreuse4") +
  ylab("Average Ozone Concentration (ppm)") +
  ggtitle("Monthly Ozone Averages (2010 - 2020)") 

# View plot
MonthlyOzone.plot
summary(MonthlyOzone.trend)
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: Based on the monthly ozone averages graph, ozone concentrations show an overall decrease over the past 10 years. The decreasing S values per season are indicative of this decreasing trend. While they do vary(dipping as low as -17 to as high as 15), I believe this is because of the variance between seasons; I think it's significant that the first S value is 15 and the last is 11, showing an overall decrease. However, the p-values for each season do not indicate significance (nothing is less than 0.05), which may just be because of the variance in between seasons. 

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

# Subtract seasonal component from monthly time series
GaringerOzone.monthly.nonseas <- GaringerOzone.monthly.ts -
  GaringerOzone.monthly.decomposed$time.series[,1]

#16

# Run non-seasonal Mann Kendall test
MonthlyOzone.nonseas.trend <- trend::mk.test(GaringerOzone.monthly.nonseas)

# View results
MonthlyOzone.nonseas.trend
# Compare
MonthlyOzone.trend

```

> Answer: In comparison to the seasonal test, the p-value from the non-seasonal test is much lower, indicating that the non-seasonal data is more statistically significant than the seasonal data. However, the z-score for the nonseasonal test is much lower than the seasonal test, indicating that the data has unusual deviation from the mean. This shows that the nonseasonal data accounts for more outliers and unusual trends than the seasonal data. 
