---
title: "Assignment 3: Data Exploration"
author: "Jaclyn Narleski"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
# Load packages
# library(tidyverse)
# library(lubridate)
# library(here)

# Check working directory
getwd()

# Upload data sets
Neonics <- read.csv(
  file = here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"), 
  stringsAsFactors = TRUE)

Litter <- read.csv(
  file = here("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"), 
  stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Pesticides often pose a threat to the environment, especially when they're overused. Runoff of pesticides into local water sources, like streams, rivers, and lakes, can harm both the toxicology of the water itself and the aquatic organisms that reside in it. Since neonicotinoids are a common pesticide, it's vital we understand their makeup and their impact on organisms.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris that falls to the ground in forests plays a key role in the carbon budgeting and nutrient cycling of forest ecosystems. Studying litter/woody debris can be informative on what kinds of nutrients an ecosystem may or may not be receiving, or have an overabundance of. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Plotting - litter and woody debris was sampled using 40 m x 40 m and 20 m x 20 m tower plots that were randomly selected areas.  
 2. Trap placement - ground and elevated traps were used in plots depending on the type of vegetation and debris to be collected. Placement of litter traps was either purposeful or randomized depending on the plot. 
 3. Sampling frequency - ground traps were sampled once per year, while elevated traps were sampled more frequently depending on their vegetation type and forest location (deciduous or evergreen).

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
# View dataset Neonics
# View(Neonics)
# Data features 4,623 entries/rows and 30 columns

# View dataset Litter
# View(Litter)
# Data features 188 entries/rows and 19 columns
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
# Summary of Neonics "Effects" column
summary(Neonics$Effect)

```

> Answer: Based on the dataset, the most common effects of neonics that are studied are population and mortality. These effects could be of particular interest since heightened mortality rates of a species due to a pesticide could be indicative of overuse/danger to a species, and population abundance could also indicate what species(s) have decreased due to environmental or epidemiological changes. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
# Help function to see how summary() works
# help("summary")

# Summary to see the six most commonly studied species using the maxsum argument
summary(Neonics$Species.Common.Name, maxsum = 7)

```

> Answer: The six most commonly studied species in this dataset are the Honey Bee, Parasitic Wasp, the Buff-Tailed Bumblebee, Carniolan Honey Bee, the Bumble Bee, and the Italian Honeybee. These species are all pollinators (the wasp less so), which would be of interest over other insects since they are helpful for keeping crop and flower populations healthy and alive.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
# View dataframe Neonics

# View(Neonics)

```

> Answer: The values in Conc.1..Author are factors. They're likely not numeric because different concentrations have different categories associated with them and therefore need to be sorted that way rather than numerically. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}

# Check to see function of geom_freqpoly()
# help("geom_freqpoly")

# Generate plot of the number of studies conducted by publication year
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year), bins = 37) # 37  years total

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}

# Generate plot of the number of studies conducted by publication year and
# different colors designate different test locations
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 37)

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the lab and the natural field. The lab test locations definitely increase over time while the natural field decreases.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}

# Generate bar graph of endpoint counts
ggplot(data = Neonics, aes(x = Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

> Answer: The two most common endpoints are NOEL and LOEL. LOEL stands for lowest-observable-effect-level, or the lowest dose/concentration reported to produced signifcant effects. NOEL stands for no-observable-effect-level, or the highest dose/concentration reported that produced no significant effects. Both of these are terrestrial endpoints.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}

# Determine class of collectDate
class(Litter$collectDate)

# Convert class of factors collectDate to characters
date_char <- as.character(Litter$collectDate)

# Convert class of characters date_char to dates
litterdate <- as.Date(date_char)

# Check class
class(litterdate)

# Check dates that were sampled in 2018
unique(litterdate)

```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

# Use unique() function to determine amount of different plots sampled at Niwot
# Ridge
unique(Litter$namedLocation)

```

> Answer: 12 different plots were sampled at Niwot Ridge. The unique() function assesses data in levels versus the summary function that performs a count. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

# Create a bar graph of functionalGroup counts
ggplot(data = Litter, aes(x = functionalGroup)) +
  geom_bar()

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}

# Create a boxplot of dryMass by functionalGroup
ggplot(Litter) +
  geom_boxplot(aes(x = dryMass, y = functionalGroup))

# Create a violin plot of dryMass by funtionalGroup
ggplot(Litter) +
  geom_violin(aes(x = dryMass, y = functionalGroup))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot is a more effective visualization because it displays the mean, median, and Q1/Q3 values, as well as outliers. This allows us to see which functional group actually had the greatest density of dry mass, while the violin plot is skewed due to outliers.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles, mixed litter, and twigs/branches
