---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Jaclyn Narleski"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

library(tidyverse)
library(agricolae)
library(lubridate)
library(corrplot)
library(here)
here()

LTER.Lake <-
  read.csv(here(
    "~/EDE_Fall2024/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))

LTER.Lake$sampledate <- as.Date(LTER.Lake$sampledate, format = "%m/%d/%Y")

#2

rosetheme <- theme(
  panel.background = element_rect(fill = "thistle1"),
  panel.border = element_rect(color = "violetred4", fill = NA, linetype = 1),
  plot.title = element_text(
    hjust = 0.5, size = 12, color = "violetred4", 
    face = "bold", family = "Lato"),
  axis.title = element_text(size = 10, color = "violetred4", family = "Lato"),
  panel.grid.major = element_line(
    color = "violetred4", linetype = 2, size = 0.25),
  panel.grid.minor = element_line(
    color = "violetred3", linetype = 3, size = 0.25),
  axis.text = element_text(color = "violetred4", family = "Lato"),
  axis.ticks = element_line(color = "violetred4"),
  legend.title = element_text(size = 8, color = "violetred4", family = "Lato"),
  legend.position = "right"
)

theme_set(rosetheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature during July does not change with depth across all lakes.
Ha: The mean lake temperature during July lowers with deeper depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

LTER.Lake.July <- LTER.Lake %>%
  filter(month(sampledate) == 7) %>%
  select(sampledate,lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5

Julytemp.by.depth <- LTER.Lake.July %>%
  ggplot(aes(
    x = temperature_C,
    y = depth
  )) +
  geom_point(color = "violetred4", size = 1, alpha = 0.75) +
  geom_smooth(method = "lm", color = "gold2", size = 1.2) +
  xlab("Temperature [C]") +
  scale_x_continuous(limits = c(0,35), expand = expansion(c(0.05,0.05))) +
  ylab("Depth (m)") +
  scale_y_continuous(limits = c(0,NA), expand = expansion(c(0.05, 0.05))) +
  ggtitle("July Temperatures by Lake Depth")

Julytemp.by.depth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggest that deeper lake depths have lower temperatures, especially when the the lake depth is below 5 meters. The distribution of the points suggests that temperatures remain more stable and warmer above 5 meters, but could possibly vary by lake since surface temperatures seem to vary between 15 C to 28 C.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

DepthTempRegression <- lm(LTER.Lake.July$temperature_C ~ LTER.Lake.July$depth)

summary(DepthTempRegression)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The linear regression shows statistical significance between lake temperature and depth, as indicated by the low p-value and t-statistic of 3.835. The R^2 value of 0.7387 suggests that almost 75% of the changes in lake temperature can be correlated with its depth. The high degrees of freedom (9726) also confirms the results and correlation between depth and lake temperature is credible. Finally, for every meter of depth, temperature is predicted to change by about 2 C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

TempAIC <- lm(data = LTER.Lake.July, temperature_C ~ depth + year4 + daynum)

step(TempAIC)

#10

TempRegression <- lm(data = LTER.Lake.July, 
                     temperature_C ~ year4 + depth)
summary(TempRegression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer:  It seems that depth and year have the best shot at predict temperature using a multiple regression, although not much better than just depth. The model explains about the same amount of observed variance as just using depth

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

LTER.anova <- aov(data = LTER.Lake.July, temperature_C ~ lakename)
summary(LTER.anova)

LTER.linear <- lm(data = LTER.Lake.July, temperature_C ~ lakename)
summary(LTER.linear)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Based on the ANOVA test, there is a significant difference in mean temperature among lakes because of the small p-value - it rejects the null hypothesis. However, the linear regression shows that the lakes that really have the most significant differences in mean temperature are East Long Lake, Hummingbird Lake, Paul Lake, Peter Lake, Tuesday Lake, and West Long Lake. While Ward Lake and Crampton Lake still have significant difference, they have higher p-values, so one could argue their differences are less so.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

Laketemp.by.depth <- LTER.Lake.July %>%
  ggplot(aes(
    x = temperature_C,
    y = depth,
    color = lakename)
  ) + 
  geom_smooth(method = "lm", se = FALSE) +
  geom_point()

Laketemp.by.depth

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
