---
title: "Assignment 5: Data Visualization"
author: "Jaclyn Narleski"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

# load packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

# verify directory
here()

# read csv files
PeterPaul.Processed <- read.csv(here(
"./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
)

NiwotRidge.Processed <- read.csv(here(
  "./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 

# change dates to date format
PeterPaul.Processed$sampledate <-
  as.Date(as.character(PeterPaul.Processed$sampledate))
NiwotRidge.Processed$collectDate <-
  as.Date(NiwotRidge.Processed$collectDate)

# check
class(PeterPaul.Processed$sampledate)
class(NiwotRidge.Processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

# build plot theme
plot.theme <- theme(
  # set title elements
  plot.title = element_text(hjust = 0.5, size = 12, 
                            color = "darkgreen", face = "bold"),
  # set x axis elements
  axis.title.x = element_text(size = 10, color = "darkgreen"),
  # set y axis elements
  axis.title.y = element_text(size = 10, color = "darkgreen"),
  # set legend title elements
  legend.title = element_text(size = 10, color = "darkgreen"),
  # set plot background
  plot.background = element_rect(color = "lightgray")
)

# set theme as default
theme_set(plot.theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

# create ggplot
phosphorous.phosphate <- PeterPaul.Processed %>%
  ggplot(aes(
    x=po4,
    y=tp_ug)
    ) +
  geom_point() +
  # split into 2 plots by lake
  facet_wrap(facets = vars(lakename), nrow = 2) +
  # set lines of best fit
  geom_smooth(method = "lm", color = "yellowgreen") +
  # hide extreme values greater than 45
  xlim(0,45) +
  ggtitle("Phosphorous to Phosphate Ratios\nin Peter & Paul Lakes") +
  ylab("Total\nPhosphorous") +
  xlab("Phosphate")

# view plot
phosphorous.phosphate

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

# change months to factorsk, set levels and abbreviations
PeterPaul.Processed$month <- factor(PeterPaul.Processed$month,
           levels = 1:12,
           labels = month.abb)

# create ggplot for temperature
peterpaul.temp <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=temperature_C,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Lake Temperatures by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Temperature [C]")
    
# view plot
peterpaul.temp

# create ggplot for total phosphorous
peterpaul.tp <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=tp_ug,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Total Phosphorous by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Total\nPhosphorous")

# view plot
peterpaul.tp

# create ggplot for total nitrogen
peterpaul.tn <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=tn_ug,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Total Nitrogen by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Total\nNitrogen")

# view plot
peterpaul.tn

# combine plots using plot_grid()
combined.peterpaul.box <- plot_grid(
  # set legend positions to none and modify title and axis titles
  peterpaul.temp + theme(legend.position = "none",
                         axis.title.x = element_blank(),
                         plot.title = element_text(size = 10)),
  peterpaul.tn + theme(legend.position = "none",
                       axis.title.x = element_blank(),
                       plot.title = element_text(size = 10)),
  peterpaul.tp + theme(legend.position = "none",
                       axis.title.x = element_blank(),
                       plot.title = element_text(size = 10)), 
  ncol = 1,
  # align vertical axis
  align = "v")

# retrieve legend and set to right side
legend <- get_legend(
  peterpaul.temp + theme(legend.position = "right")
)

# manually combine legend and combined box plots for final product
final.plot <- plot_grid(combined.peterpaul.box, legend, ncol = 2,
                        # adjust margins
                        rel_widths = c(0.8, 0.2))

# view plot
final.plot

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: During warmer months where lake temperatures are higher, there tends to be greater amount of high phosphorous and nitrogen measurements, especially in Peter Lake. These measurements fall as the lake temperatures lower.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

# create ggplot
needle.litter <- NiwotRidge.Processed %>%
  # filter for needle litter only
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(
    x=collectDate,
    y=dryMass,
    color=nlcdClass)
  ) +
  geom_point() +
  ggtitle("Dry Mass of Needle Litter") +
  xlab("Date") +
  ylab("Dry Mass") +
  labs(color = "NLCD Class")

# view plot
needle.litter

#7

# create ggplot
needle.litter.byclass <- NiwotRidge.Processed %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(aes(
    x=collectDate,
    y=dryMass)
  ) +
  geom_point() +
  # split into 3 plots by NLCD Class
  facet_wrap(facets = vars(nlcdClass), nrow = 3) +
  ggtitle("Dry Mass of Needle Litter") +
  xlab("Date") +
  ylab("Dry Mass")

# view plots
needle.litter.byclass

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective since one is able to see the data for each NLCD class clearer when it's separated and how these measurements change over time. In one plot, we are only able to see outliers clearly. 