---
title: "Assignment 5: Data Visualization"
author: "Jaclyn Narleski"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

# load packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

# verify directory
here()

# read csv files
PeterPaul.Processed <- read.csv(here(
"./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
)

NiwotRidge.Processed <- read.csv(here(
  "./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 

# change dates to date format
PeterPaul.Processed$sampledate <-
  as.Date(as.character(PeterPaul.Processed$sampledate))
NiwotRidge.Processed$collectDate <-
  as.Date(NiwotRidge.Processed$collectDate)

# check
class(PeterPaul.Processed$sampledate)
class(NiwotRidge.Processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

plot.theme <- theme(
  plot.title = element_text(hjust = 0.5, size = 12, 
                            color = "darkgreen", face = "bold"),
  axis.title.x = element_text(size = 10, color = "darkgreen"),
  axis.title.y = element_text(size = 10, color = "darkgreen"),
  legend.title = element_text(size = 10, color = "darkgreen"),
  plot.background = element_rect(color = "lightgray")
)

theme_set(plot.theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

phosphorous.phosphate <- PeterPaul.Processed %>%
  ggplot(aes(
    x=po4,
    y=tp_ug)
    ) +
  geom_point() +
  facet_wrap(facets = vars(lakename), nrow = 2) +
  geom_smooth(method = "lm", color = "yellowgreen") +
  xlim(0,45) +
  ggtitle("Phosphorous to Phosphate Ratios\nin Peter & Paul Lakes") +
  ylab("Total\nPhosphorous") +
  xlab("Phosphate")

phosphorous.phosphate

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

PeterPaul.Processed$month <- factor(PeterPaul.Processed$month,
           levels = 1:12,
           labels = month.abb)

peterpaul.temp <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=temperature_C,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Lake Temperatures by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Temperature [C]")
    

peterpaul.temp

peterpaul.tp <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=tp_ug,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Total Phosphorous by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Total\nPhosphorous")

peterpaul.tp

peterpaul.tn <- PeterPaul.Processed %>%
  ggplot(aes(
    x=month,
    y=tn_ug,
    color=lakename)
    ) +
  geom_boxplot() +
  ggtitle("Total Nitrogen by Month") +
  labs(color = "Lake") +
  xlab("Month") +
  ylab("Total\nNitrogen")

peterpaul.tn

combined.peterpaul.box <- plot_grid(
  peterpaul.temp, peterpaul.tn, peterpaul.tp) 
#  axis.title.x = element_blank() +
#  legend.position = "none"

combined.peterpaul.box
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: