---
title: "Assignment 10: Data Scraping"
author: "Jaclyn Narleski"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 

# Load packages
library(tidyverse)
library(rvest)
library(lubridate)
library(dataRetrieval)
library(here)

# Check working directory
here()

# Create personal theme
water_theme <- theme(
  plot.title = element_text(color = "darkblue", face = "bold", hjust = 0.5),
  axis.title.x = element_text(color = "darkblue"),
  axis.title.y = element_text(color = "darkblue", angle = 90, hjust = 0.5),
  panel.background = element_rect(fill = "lightblue1")
)

# Set my theme
theme_set(water_theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2023 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

# Read in web address
LWSP <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023')

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

# Scrape data using nodes
WaterSystem <- LWSP %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
  html_text()
PWSID <- LWSP %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)') %>%
  html_text()
Ownership <- LWSP %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()

# Read whole MGD table sections
MGD <- LWSP %>% 
  html_nodes('th~ td+ td') %>%
  html_text()
```

4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2023, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

# Set month vector in order of months from left to right on website
Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", 
           "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")

# Repeat year for lenght of months
Year <- rep(2023, length(Month))

# Paste year and month with 1st day of month
Date <- paste0(Year, Month, "1")

# Format into date object
Date <- as.Date(Date, format = "%Y %b %d")

# Generate data frame
DurhamWater <- data.frame(
  "Date" = Date,
  "Water System" = WaterSystem,
  "PWSID" = PWSID,
  "Ownership" = Ownership,
  "Max Day Use" = as.numeric(MGD)
 )

#5 

# Create ggplot
DurhamWaterPlot <- DurhamWater %>%
  ggplot(aes(
    x = Date,
    y = Max.Day.Use)
  ) +
  geom_line() +
  ggtitle("Maximum Daily Water Use") +
  ylab("Maximum Daily Use") 

# View plot
DurhamWaterPlot
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data, returning a dataframe.
**Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 

# Scrape it function
scrape.it <- function(PWSID, Year){
  the_url <- paste0(
    "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
    PWSID,"&year=",Year) # Paste PWSID and Year inputs into URL
  the_website <- read_html(the_url) # Read url

# Scrape nodes    
WaterSystem <- the_website %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)') %>% 
  html_text()
PWSID_scraped <- rep(PWSID, length(WaterSystem))
Ownership <- the_website %>%
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)') %>%
  html_text()
MGD <- the_website %>% 
  html_nodes('th~ td+ td') %>%
  html_text()

# Set month vector, repeat year for length of months, and set to date object
Month <- c("Jan", "May", "Sep", "Feb", "Jun", "Oct", 
           "Mar", "Jul", "Nov", "Apr", "Aug", "Dec")
Year1 <- rep(Year, length(Month))
Date <- paste0(Year1, Month, "1")
Date <- as.Date(Date, format = "%Y %b %d")

# Generate data frame
water_data <- data.frame(
  "Date" = Date,
  "Water System" = WaterSystem,
  "PWSID" = PWSID_scraped,
  "Ownership" = Ownership,
  "Max Day Use" = as.numeric(MGD)
)

return(water_data)

}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 

# Scrape 2015 data
DurhamWater2015 <- scrape.it('03-32-010',2015)

# Plot data
DurhamWater2015plot <- DurhamWater2015 %>%
  ggplot(aes(
    x = Date,
    y = Max.Day.Use
  )) +
  geom_line() +
  ggtitle("Maximum Daily Water Use (Durham, 2015)") +
  ylab("Maximum Daily Use") 

# View plot
DurhamWater2015plot

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 

# Scrape 2015 Asheville data
AshevilleWater2015 <- scrape.it('01-11-010', 2015)

# Combine data into one data frame 
CombinedData <- rbind(DurhamWater2015, AshevilleWater2015)

# Generate plot
Asheville_Durham_Compare <- CombinedData %>%
  ggplot(aes(
    x = Date,
    y = Max.Day.Use,
    color = Water.System
  )) +
  geom_line() +
  ggtitle("Maximum Daily Water Use (2015)") +
  ylab("Maximum Daily Use") +
  labs(color = "City Water System")

# View plot
Asheville_Durham_Compare

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2022.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

# Set object for multi-year vector 
Multiyears <- seq(2018,2022)

# Use map2 to run over scrape it function with 2 inputs
AshevilleWater_18_22 <- map2(
  rep('01-11-010', length(Multiyears)), Multiyears, scrape.it) %>%
  bind_rows()

# Generate plot
Asheville_multi <- AshevilleWater_18_22 %>%
  ggplot(aes(
    x = Date,
    y = Max.Day.Use)
  ) +
  geom_line() +
  geom_smooth(method = 'loess') +
  scale_x_date(date_breaks = '1 year', date_labels = '%Y') +
  ggtitle("Maximum Daily Water Use (2018-2022)") +
  ylab("Maximum Daily Use") 

# View plot
Asheville_multi

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Overall, based off the geom_smooth line, Asheville does display an upward trend in water usage oer time.
>
